Categorical Foundations of E-Graphs With Binding

### 2025-07-11 POPL2026/paper/CategoricalFoundationsOfEGraphsWithBinding.tex
touch up conclusion, fix broken references, move empirical analysis to appendix

\section{Conclusion}
In this work we introduced e-graphs with bindings formalised as morphisms of a closed symmetric monoidal $\catname{SLat}$-category.
We then provided a combinatorial representation for such morphisms in terms of hierarchical hypergraphs (e-hypergraphs) with a suitable notion of sound and complete DPO rewriting, making it possible for an e-hypergraph to represent equivalences derivable from a set of equations $\mathcal{E}$.
An important future work is exploring whether such combinatorial structures may lead to better algorithmics for doing equality saturation for $\lambda$-calculus.

>>>>

\section{Conclusion}
In this work we introduced e-graphs with bindings formalised as morphisms of a closed symmetric monoidal $\catname{SLat}$-category.
In doing so, we are able to leverage its internal language to provide variable binding in $\lambda$-calculus-like languages by metatheoretic $\lambda$-abstraction.
We then provided a combinatorial representation for such morphisms in terms of closed e-hypergraphs with a suitable notion of sound and complete DPO rewriting, making it possible for a closed e-hypergraph to represent equivalences derivable from a set of equations $\mathcal{E}$.
We argued for the utility of our techniques in the representation of programming languages (which inherently require variable binding) as e-graphs, contrasting with the more general-purpose slotted e-graphs, and advocating for the benefits of string-diagrammatic syntax.
In future, we hope to implement an efficient version of e-graphs for hypergraphs as a data structure using DPO rewriting as its foundation to implement e-graph operations, with the theory we have developed here acting as a roadmap.

### 2025-07-11 POPL2026/paper/sections/appendix.tex
touch up conclusion, fix broken references, move empirical analysis to appendix

>>>>

\section{Modelling $\lambda_{\text{lsub}}$-calculus in slotted e-graphs}%
\label{sec:e-graph-explicit-subst}

\begin{table}
	\begin{tabular}{lcc}
		\texttt{beta}                  & $(\lambda x . t) u = t[x / u]$                       &                                                                    \\
		\texttt{subst\_same\_var}      & $x[x / u] = u$                                       &                                                                    \\
		\texttt{subst\_diff\_var}      & $y[x / u] = y$                                       & if $x \not \in \mathcal{F}(u)$                                     \\
		\texttt{subst\_application}    & $(uu')[x / v] = u[x / v]u'[x / v]$                   &                                                                    \\
		\texttt{subst\_lambda}         & $(\lambda x . u)[y / v] = \lambda x . (u[y / v])$    & if $x \not \in \mathcal{F}(v)$                                     \\
		\texttt{subst\_interchange\_1} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v[x' / v']]$ & if $x' \in \mathcal{F}(v)$ and $x \in \mathcal{F}(v')$             \\
		\texttt{subst\_interchange\_2} & $(u[x / v])[x' / v'] = (u[x / v[x' / v']])$          & if $x' \not \in \mathcal{F}(u)$                                    \\
		\texttt{subst\_comm}           & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v]$          & if $x \not \in \mathcal{F}(v')$ and $x' \not \in \mathcal{F}(v)~.$ \\
	\end{tabular}
	\caption{Encoding of string diagram equivalence and explicit substitution for $\lambda_{\text{lsub}}$}
	\label{tbl:e-graph-subst-rules}
\end{table}

Consider the equational presentation of string diagram equivalence and the $\lambda_{\text{lsub}}$ reduction rules given by~\autoref{tbl:e-graph-subst-rules}.
These can be encoded in the framework of slotted e-graphs.
To demonstrate the effect that a particular set of rewrites has on the number of e-nodes of an e-graph, we saturate the lambda term $\text{plus}(2, 2)$ where $\text{plus}$ and $2$ are encoded using Church encoding.
The numbers of e-nodes after the e-graph is saturated for each set of rewrite rules are presented in~\autoref{tbl:slotted}.

\begin{table}
	\begin{tabular}{lcccc}
		Set of equations & explicit subst & explicit subst & explicit subst & explicit subst \\
		                 &                & + comm         & + interchange  & + interchange  \\
		                 &                &                &                & + comm         \\
		\# of e-nodes    & 161            & 288            & 289            & 356
	\end{tabular}
	\caption{Slotted e-graphs benchmarks}
	\label{tbl:slotted}
\end{table}

The \enquote{explicit subst} set of rewrite rules consist of the first five rules of~\autoref{tbl:e-graph-subst-rules}.
We can see that additionally saturating the e-graph with respect to interchange and commutativity rules increases the number of e-nodes by a factor of 2.
These are the exact equations that are absorbed by using string-diagrammatic representation of $\lambda_{\text{lsub}}$-terms.
Notably, string diagrams (and (closed e-)hypergraphs) do not absorb the equation \texttt{subst\_lambda} but we believe it is practically feasible to do rewriting modulo this rule by traversing through the boxes when searching for a redex.
Furthermore, (cospans of) closed e-hypergraphs, as presented in~\autoref{sec:closed-e-hyp}, do not absorb the equations for associativity, commutativity and unitality for $\comonoid$ and $\counit$, but the constraints on the cospans and the boundary complement may be relaxed to absorb those as well by following the work of~\cite{zanassi_comonoid} with which our approach is completely compatible.

### 2025-07-11 POPL2026/paper/sections/categorical_semantics.tex
touch up conclusion, fix broken references, move empirical analysis to appendix

To aid reasoning, we introduce a new language of string diagrams for closed symmetric monoidal $\catname{SLat}$-categories, using a hierarchical \enquote{dashed box} structure to capture the join operation on morphisms (as was done by~\citet{ghica2024equivalencehypergraphsegraphsmonoidal}) as well as another hierarchical \enquote{rounded box} for $\lambda$-abstraction.
\autoref{fig:egraph-strings} displays the generators of this language, which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for $\lambda$-abstraction as established by~\citet{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice enrichment as introduced by~\citet{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.

>>>>

To aid reasoning, we introduce a new language of string diagrams for closed symmetric monoidal $\catname{SLat}$-categories, using a hierarchical \enquote{dashed box} structure to capture the join operation on morphisms (as was done by~\citet{tiurin2025equivalencehypergraphsdporewriting}) as well as another hierarchical \enquote{rounded box} for $\lambda$-abstraction.
\autoref{fig:egraph-strings} displays the generators of this language, which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for $\lambda$-abstraction as established by~\citet{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice enrichment as introduced by~\citet{tiurin2025equivalencehypergraphsdporewriting}; the first component denotes an empty diagram.

### 2025-07-11 POPL2026/paper/sections/introduction.tex
touch up conclusion, fix broken references, move empirical analysis to appendix

E-graphs (short for \emph{equality graphs})~\cite{EggPaper} are an efficient data structure for non-destructive rewriting, incorporating equivalence representation, used in automated theorem proving, compiler optimisation, and program analysis.
They mitigate the \emph{phase-ordering problem} of term rewriting: the way rewrites are scheduled may enable or block further rewrites due to introduction or elimination or redexes, resulting in a suboptimal result.
This is achieved by making rewrites on e-graphs non-destructive: an application of a rewrite rule, instead of replacing the left-hand side by the right-hand side, adds the latter to the e-graph so that all pre-existing redexes are protected and new ones are introduced.
Explosive growth of the e-graph is avoided by the clever use of sharing of vertices.

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest, particularly for optimisation and synthesis using the technique of \emph{equality saturation}~\cite{10.1145/1594834.1480915, flatt2022_small, EggPaper,flatt_small_2022}, wherein rewrites to an e-graph are repeated until a fixpoint (or timeout) is reached.
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
We are interested in using e-graphs in the context of functional programming languages, which builds upon an internal representation of the $\lambda$-calculus.
Pioneering work in this direction has been done by~\citet{koehler2022sketchguided}, where the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.

@@

First, let us compare how substitution works in conventional e-graphs as compared to our proposed approach.

>>>>

E-graphs (short for \emph{equality graphs})~\cite{EggPaper} are an efficient data structure for non-destructive rewriting, incorporating equivalence representation, used in automated theorem proving, compiler optimisation, and program analysis.
They mitigate the \emph{phase-ordering problem} of term rewriting: the way rewrites are scheduled may enable or block further rewrites due to introduction or elimination or redexes, resulting in a suboptimal result.
This is achieved by making rewrites on e-graphs non-destructive: an application of a rewrite rule, instead of replacing the left-hand side by the right-hand side, adds the latter to the e-graph so that all pre-existing redexes are protected and new ones are introduced.
Exponential growth of the e-graph is avoided by the clever use of sharing of vertices.

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest, particularly for optimisation and synthesis using the technique of \emph{equality saturation}~\cite{10.1145/1594834.1480915, flatt_small_2022, EggPaper,flatt_small_2022}, wherein rewrites to an e-graph are repeated until a fixpoint (or timeout) is reached.
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
We are interested in using e-graphs in the context of functional programming languages, which builds upon an internal representation of the $\lambda$-calculus.
Pioneering work in this direction has been done by~\citet{koehler2022sketchguided}, where the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.

@@

First, let us compare how substitution works in conventional e-graphs, which lack specialised support for binding, as compared to our proposed approach.

### 2025-07-11 POPL2026/paper/sections/lambda-calculus.tex
touch up conclusion, fix broken references, move empirical analysis to appendix

This calculus is explicitly designed to capture the representation of terms as proof nets: for two terms $t, u$ in the $\lambda_{\text{lsub}}$-calculus, $t \sim u$ if and only if both map to the same proof net~\cite{?}.
By further extending the equivalence relation $\sim$, taking the symmetric closure by $\to_{ls}$ and $\to_{gc}$ and thus identifying terms under elimination of explicit substitution, we determine a \emph{string diagram equivalence} $\approx$ of $\lambda_{\text{lsub}}$-terms, determining a $\lambda^\approx_{\text{lsub}}$-calculus.
Predictably, $t \approx u$ if and only if both map to the same string diagram~\cite[§ 5.1.1]{ghica2024stringdiagramslambdacalculifunctional}; this allows us to exploit the string diagram setting of e-hypergraphs to absorb many\footnote{The reason this is not \emph{all} reductions is because e-hypergraphs quotient away only the \emph{structural} isomorphisms of symmetric monoidal closed $\catname{SLat}$-categories (e.g.\ associators, unitors, etc.) which generate the \enquote{isotopy of string diagrams}; even in the graphical equivalence relation $\sim$, while~\autoref{eq:subst-lambda} is validated in every symmetric monoidal closed $\catname{SLat}$-category, it is not so by isotopy, but rather the adjunction $ - \otimes A \dashv A \multimap -$. In other words, terms floating freely through the boundaries of rounded boxes is validated by our semantics, but produces concretely different e-hypergraphs --- this is the purpose of rectification by quotienting along $\mathcal{S}$ in~\autoref{prop:quotient-structural}, and manifests as additional DPO rewrites to be imposed.} of the reductions arising due to the equivalence $\approx$ into our representation rather than explicitly providing them as equations attached to a closed symmetric monoidal theory presenting the $\lambda^\approx_{\text{lsub}}$-calculus.

@@

\autoref{fig:plus-1-2-programs} shows some programs which are semantically equivalent but syntactically distinct.
As alluded to in~\autoref{sec:vs-e-graphs-with-binding}, the traditional (slotted) e-graph approach to modelling the equivalence of these terms (\autoref{fig:plus-1-2-terms}) relies on explicitly encoding substitution operations as e-graph nodes, and their behaviour (the string diagram equivalence relation $\approx$) as rewrite rules, as in~\autoref{fig:e-graph-substitution} --- not only is this more complicated, but it also leads to a blow-up in the state space required to perform equality saturation in any extension of the $\lambda^\approx_{\text{lsub}}$-calculus, because we are really rewriting $\lambda_{\text{lsub}}$-terms and computational resources are spent rewriting between equivalent terms.
\textcolor{cyan}{To illustrate this, let's consider the following equations induced by the graphical equivalence and the $\lambda_{\text{lsub}}$ reduction rules}

\begin{tabular}{lcc}
	\texttt{beta}                  & $(\lambda x . t) u = t[x / u]$                       &                                                                    \\
	\texttt{subst\_same\_var}      & $x[x / u] = u$                                       &                                                                    \\
	\texttt{subst\_diff\_var}      & $y[x / u] = y$                                       & if $x \not \in \mathcal{F}(u)$                                     \\
	\texttt{subst\_application}    & $(uu')[x / v] = u[x / v]u'[x / v]$                   &                                                                    \\
	\texttt{subst\_lambda}         & $(\lambda x . u)[y / v] = \lambda x . (u[y / v])$    & if $x \not \in \mathcal{F}(v)$                                     \\
	\texttt{subst\_interchange\_1} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v[x' / v']]$ & if $x' \in \mathcal{F}(v)$ and $x \in \mathcal{F}(v')$             \\
	\texttt{subst\_interchange\_2} & $(u[x / v])[x' / v'] = (u[x / v[x' / v']])$          & if $x' \not \in \mathcal{F}(u)$                                    \\
	\texttt{subst\_comm}           & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v]$          & if $x \not \in \mathcal{F}(v')$ and $x' \not \in \mathcal{F}(v)~.$ \\
\end{tabular}

\textcolor{cyan}{These can be encoded in a framework of slotted e-graphs.
	To demonstrate the effect that a particular set of rewrites has on the number of e-nodes of an e-graph, we saturate the lambda term $plus 2 2$ where $plus$ and $2$ are encoded using Church encoding.
	The numbers of e-nodes after the e-graph is saturated for each set of rewrite rules are presented in \autoref{tbl:slotted}.
}

\begin{table}
	\begin{tabular}{lcccc}
		Set of equations & explicit subst & explicit subst & explicit subst & explicit subst \\
		                 &                & + comm         & + interchange  & + interchange  \\
		                 &                &                &                & + comm         \\
		\# of e-nodes    & 161            & 288            & 289            & 356
	\end{tabular}
	\caption{Slotted e-graphs benchmarks}
	\label{tbl:slotted}
\end{table}

\textcolor{cyan}{The \texttt{explicit subst} set of rewrite rules consist of the first five rules above.
	We can see that additionally saturating the e-graph with respect to interchange and commutativity rules increases the number of e-nodes by a factor of 2.
	These are the exact equations that are absorbed by using string-diagrammatic representation of $\lambda$-terms.
	Notably, string diagrams (and (e-)hypergraphs) do not absorb the equation \texttt{subst\_lambda} but we believe it is practically feasible to do rewriting modulo this rule by traversing through the boxes when searching for a redex.
	Furthermore, (cospans of) e-hypergraphs, as presented in \autoref{sec:closed-e-hyp}, do not absorb the equations for associativity, commutativity and unitality for $\comonoid$ and $\counit$, but the constraints on the cospans and the boundary complement may be relaxed to absorb those as well by following the work of~\cite{zanassi_comonoid} with which our approach is completely compatible.}

In contrast, in our framework, $\lambda^\approx_{\text{lsub}}$-terms are directly modelled as string diagrams (\autoref{fig:plus-1-2-string-diagram}\footnote{Note that dashed boxes are elided for convenience here because no rewriting has occurred.}).
We do however note that the conventional approach is more flexible, in the sense that it may model more theories with binders (e.g.\ $\pi$-calculus), whereas our approach is essentially specialised to $\lambda$-calculus, relying heavily on the metatheoretic internal language of the ambient setting.

>>>>

This calculus is explicitly designed to capture the representation of terms as proof nets: for two terms $t, u$ in the $\lambda_{\text{lsub}}$-calculus, $t \sim u$ if and only if both map to the same proof net~\cite{accattoliJumpingBoxGraphical2011}.
By further extending the equivalence relation $\sim$, taking the symmetric closure by $\to_{ls}$ and $\to_{gc}$ and thus identifying terms under elimination of explicit substitution, we determine a \emph{string diagram equivalence} $\approx$ of $\lambda_{\text{lsub}}$-terms, determining a $\lambda^\approx_{\text{lsub}}$-calculus.
Predictably, $t \approx u$ if and only if both map to the same string diagram~\cite[§ 5.1.1]{ghica2024stringdiagramslambdacalculifunctional}; this allows us to exploit the string diagram setting of e-hypergraphs to absorb many\footnote{The reason this is not \emph{all} reductions is because e-hypergraphs quotient away only the \emph{structural} isomorphisms of symmetric monoidal closed $\catname{SLat}$-categories (e.g.\ associators, unitors, etc.) which generate the \enquote{isotopy of string diagrams}; even in the graphical equivalence relation $\sim$, while~\autoref{eq:subst-lambda} is validated in every symmetric monoidal closed $\catname{SLat}$-category, it is not so by isotopy, but rather the adjunction $ - \otimes A \dashv A \multimap -$. In other words, terms floating freely through the boundaries of rounded boxes is validated by our semantics, but produces concretely different e-hypergraphs --- this is the purpose of rectification by quotienting along $\mathcal{S}$ in~\autoref{prop:quotient-structural}, and manifests as additional DPO rewrites to be imposed.} of the reductions arising due to the equivalence $\approx$ into our representation rather than explicitly providing them as equations attached to a closed symmetric monoidal theory presenting the $\lambda^\approx_{\text{lsub}}$-calculus.

@@

\autoref{fig:plus-1-2-programs} shows some programs which are semantically equivalent but syntactically distinct.
As alluded to in~\autoref{sec:vs-e-graphs-with-binding}, the traditional (slotted) e-graph approach to modelling the equivalence of these terms (\autoref{fig:plus-1-2-terms}) relies on explicitly encoding substitution operations as e-graph nodes, and their behaviour (the string diagram equivalence relation $\approx$) as rewrite rules, as in~\autoref{fig:e-graph-substitution} --- not only is this more complicated, but it also leads to a blow-up in the state space required to perform equality saturation in any extension of the $\lambda^\approx_{\text{lsub}}$-calculus, because we are really rewriting $\lambda_{\text{lsub}}$-terms and computational resources are spent rewriting between equivalent terms (an empirical analysis is provided in~\autoref{sec:e-graph-explicit-subst}).
In contrast, in our framework, $\lambda^\approx_{\text{lsub}}$-terms are directly modelled as string diagrams (\autoref{fig:plus-1-2-string-diagram}\footnote{Note that dashed boxes are elided for convenience here because no rewriting has occurred.}).
We do however note that the slotted approach is more flexible, in the sense that it may model more theories with binders (e.g.\ $\pi$-calculus), whereas our approach is essentially specialised to $\lambda$-calculus, relying heavily on the metatheoretic internal language of the ambient setting.
